---
title: "478 final project"
author: "Esmée Q. Kuiper and Cody Quiroz"
date: "2022-12-13"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
options(warn=-1)
```

```{r load packages, include=FALSE}
library(tidyverse)
library(dplyr)
library(lubridate)
library(viridisLite)
library(knitr)
```

### Setting up our data analysis

*First, we read in and clean our data, so that we can more easily analyze it.*

```{r load in and clean data}
HABS_data <- read_csv("478_LakeErie_data.csv") %>% 
  mutate(month = months(as.Date(Date, format = "%m/%d/%Y"))) %>% 
  mutate(year = year(as.Date(Date, format = "%m/%d/%Y"))+2000) %>% 
  mutate(nit_M = nitrate_nitrite_mgL/(62.0049*1000), 
                 total_dissolved_posphorus_M = total_dissolved_posphorus_ugL/(94.9714*1000*1000),
                 DOC_M = DOC_mgL/(137.11*1000)) %>% 
  mutate(date_tidy = as.Date(Date, format = "%m/%d/%y")) 
```

*Next, we tell R that we want our months to be in chronological order (rather than alphabetical).*

```{r months ordering}
HABS_data_factored <- HABS_data %>% 
  mutate(month = factor(month, levels = c("March", "April", "May", "June",
                        "July", "August", "September", "October", "November"))) %>% 
  arrange(month)
```

### Plotting phosphorus data

```{r plot tdp data}
HABS_data %>%
  group_by(year) %>% 
  filter(total_dissolved_posphorus_M > 0, nit_M > 0, DOC_M > 0) %>%
  ggplot(aes(x = date_tidy,
             y = total_dissolved_posphorus_M, 
             color = year)) +
    geom_smooth(col = "black") +
  geom_point(alpha=0.5) + 
  scale_color_viridis_b(option = "cividis") + 
  theme_bw() + 
  labs(title = "Total Dissolved Phosphorus vs. Time") +
  xlab("Year") + 
  ylab("Total Dissolved Phosphorus (M)")

HABS_data_factored %>% 
  filter(total_dissolved_posphorus_M > 0, nit_M > 0, DOC_M > 0,
         month == c("June","July","September")) %>% 
  ggplot(aes(x = as.factor(month), 
             y = total_dissolved_posphorus_M,
             color = year)) + 
  geom_boxplot() +
  geom_point(position = position_jitterdodge(jitter.width = 0.2), 
             alpha = 0.5) +
  scale_color_viridis_b(option = "cividis") + 
  theme_bw() + 
  labs(title = "Total Dissolved Phosphorus vs. Time") +
  xlab("Month") + 
  ylab("Total Dissolved Phosphorus (M)") +
  ylim(0, 1.5e-6)

ggsave("tdp_month_plot.png",
       height = 4,
       width = 6)
```

#### Stastical analysis of phosphorus 

```{r statistical analysis of P}
phosphorus_anova <- 
  aov(total_dissolved_posphorus_M ~ month, data = HABS_data_factored)

summary(phosphorus_anova)

phosphorus_tukey <- TukeyHSD(phosphorus_anova)

phosphorus_tukey
```

### Plotting nitrate-nitrite data

```{r plot nitrate-nitrite data}
HABS_data %>%
  group_by(year) %>% 
  filter(nit_M > 0, nit_M > 0, DOC_M > 0) %>%
  ggplot(aes(x = date_tidy,
             y = nit_M, 
             color = year)) +
    geom_smooth(col = "black") +
  geom_point(alpha = 0.5) + 
  scale_color_viridis_b(option = "cividis") + 
  theme_bw() + 
  labs(title = "Total Nitrate-Nitrite vs. Time") +
  xlab("Year") + 
  ylab("Total Nitrate-Nitrite (M)")

HABS_data_factored %>% 
  filter(nit_M > 0, nit_M > 0, DOC_M > 0,
         month == c("June","July","September")) %>% 
  ggplot(aes(x = as.factor(month), 
             y = nit_M,
             color = year)) + 
  geom_boxplot() +
  geom_point(position = position_jitterdodge(jitter.width = 0.2), 
             alpha = 0.5) +
  scale_color_viridis_b(option = "cividis") + 
  theme_bw() + 
  labs(title = "Total Nitrate-Nitrite vs. Time") +
  xlab("Month") + 
  ylab("Total Nitrate-Nitrite (M)") +
  ylim(0, 1.5e-4)

#save boxplot
ggsave("nit_month_plot.png",
       height = 4,
       width = 6)
```

#### Statistical analysis of nitrate-nitrite

```{r statistical analysis of nitrate-nitrite}
nitrate_anova <- 
  aov(nit_M ~ month, data = HABS_data_factored)

summary(nitrate_anova)

TukeyHSD(nitrate_anova)
```

### Plotting dissolved organic carbon (DOC) data

```{r plot DOC data}
HABS_data %>%
  group_by(year) %>% 
  filter(nit_M > 0, nit_M > 0, DOC_M > 0) %>%
  ggplot(aes(x = date_tidy,
             y = DOC_M, 
             color = year)) +
    geom_smooth(col = "black") +
  geom_point(alpha = 0.5) + 
  scale_color_viridis_b(option = "cividis") + 
  theme_bw() + 
  labs(title = "Total DOC vs. Time") +
  xlab("Year") + 
  ylab("Total DOC (M)")

HABS_data_factored %>% 
  filter(total_dissolved_posphorus_M > 0, nit_M > 0, DOC_M > 0,
         month == c("June","July","September")) %>% 
  ggplot(aes(x = as.factor(month), 
             y = DOC_M,
             color = year)) + 
  geom_boxplot() +
  geom_point(position = position_jitterdodge(jitter.width = 0.2), 
             alpha = 0.5) +
  scale_color_viridis_b(option = "cividis") + 
  theme_bw() + 
  labs(title = "Total DOC vs. Time") +
  xlab("Month") + 
  ylab("Total DOC (M)") +
  ylim(0, 6e-5)


ggsave("DOC_month_plot.png",
       height = 4,
       width = 6)
```

#### Statistical analysis of DOC

```{r statistical analysis of DOC}
DOC_anova <- 
  aov(DOC_M ~ month, data = HABS_data_factored)

summary(DOC_anova)

TukeyHSD(DOC_anova)
```

### Preparing our data for ChemEQL

Next, we summarize key measures by month. We will use these values to create a matrix in ChemEQL.  

```{r summarize by month}
#create new data frame
HABS_data_month <- HABS_data %>% 
  group_by(month) %>% 
  filter(total_dissolved_posphorus_M > 0, nit_M > 0, DOC_M > 0) %>% 
  summarize(mean_tdp_M = mean(total_dissolved_posphorus_M), 
            mean_nit_M = mean(nit_M), 
            mean_DOC_M = mean(DOC_M))

#create data table
HABS_data_month_factored <- HABS_data_month %>% 
  mutate(month = factor(month, levels = c("March", "April", "May", "June",
                        "July", "August", "September", "October", "November"))) %>% 
  arrange(month)
kable(HABS_data_month_factored, col.names = 
        c("Month", "Mean Total Dissolved Phosphorus (M)", "Mean Nitrate-Nitrite (M)", "Mean DOC (M)"))
```
